Global Life Expectancy: Unraveling Health and Economic Determinants (2015-2020)

### 2024-12-04 paper/paper.qmd
momo

---
title: "Global Life Expectancy: Unraveling Health and Economic Determinants (2015-2020)"
subtitle: "Multiple Linear Regression Analyzing Critical Factors Shaping Life Expectancy "
author: 
  - Yanfei Huang
thanks: "Code and data are available at: [https://github.com/wendyhuan/lifeexpectancy](https://github.com/wendyhuan/lifeexpectancy)."
date: 02 December 2024
date-format: long
abstract: "This paper analyze life expectancy and its determinants through 184 countries and make the prediction of people from different Income Group. Multiple Linear regression is used to deploying the life expectanc with gender, income and region. Predictions of life expectancy of people from different income group is made according to these essential predictors. The finding indicates that the life expectancy is tend to get higher as the economic class grows. And we predict that the average age of person from low, lower-middle, upper-middle and high are 62.35, 68.40, 73.38, 79.38. The result of this paper will help in suggesting a country which area should be given importance in order to efficiently improve the life expectancy of its population."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

# Plot the percentage distribution
ggplot(life_expectancy_bins, aes(x = AgeRange, y = Percentage, fill = AgeRange)) +
  geom_bar(stat = "identity", color = "white", alpha = 0.8) +
  geom_text(aes(label = paste0(round(Percentage, 2), "%")), vjust = -0.5, size = 3.5) +
  labs(
    title = "Distribution of Life Expectancy by Age Range",
    x = "Life Expectancy (5-Year Ranges)",
    y = "Percentage"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

@@

# Create a model summary with median absolute deviation (mad) for statistics
modelsummary(
  list("Gaussian(Normal)" = model_developed),
  statistic = "mad",  # Median absolute deviation
  fmt = 2,     # Format results to 2 decimal places
  output = "kableExtra"
) %>%
  kable_styling(font_size = 8.5,
                full_width = FALSE,  # Keeps the table from spanning the full width
    position = "center", # Centers the table
    bootstrap_options = c("striped", "condensed") # Makes rows tighter
  )

@@

The income group column and the region column was not given in the raw dataset. We created a new variable called "Income_Group" by mapping over country name to four income group (lower_income, lower_middle, upper_middle, high_income) according to the index given by World Bank Income Group. This is a key variable in predicting life expectancy since it is believed that life expectancy is correlated to the income situation. We as well made a mapping to the countries according to its continent and saved as "Region". This is the variable indicates the geographic context to the analysis. Countries are mapped into six continents ("Africa", "Asia", "Europe", "North America", "South America", "Oceania") one by one.

@@

```{r}
#| label: tbl-cleaned_average
#| tbl-cap: "Cleaned data of average life expectancy of 6 years by different countries. Full cleaned Average life expectancy by country by gender could be find in https://github.com/wendyhuan/lifeexpectancy/blob/main/data/02-analysis_data/average_life_expectancy.csv"
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Create a nicely formatted kable table
bothg_average_life_expectancy %>%
  kable(
    col.names = c("Country", "Income Group", "Region", "Average Life Expectancy"),
    align = c("l", "l", "l", "l", "r"),
    digits = 2,
    output = "kableExtra"
) %>%
  kable_styling(font_size = 8.5,
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "condensed")
  ) 
```

@@

# Format the table using kable
results_df %>%
  head(100) %>%
  kable(col.names = c("Country", "Actual Life Expectancy", "Predicted Life Expectancy", "RMSE (Test)", "RMSE (Train)", "Indication"),
  align = c("l", "r", "r", "r", "r", "l"),
  digits = 2,
  output = "kableExtra"
) |>
  kable_styling(font_size = 7.5,
                full_width = FALSE,  # Keeps the table from spanning the full width
    position = "center", # Centers the table
    bootstrap_options = c("striped", "condensed"))

>>>>

---
title: "Global Life Expectancy: Unraveling Health and Economic Determinants (2015-2020)"
subtitle: "Multiple Linear Regression Analyzing Critical Factors Shaping Life Expectancy "
author: 
  - Yanfei Huang
thanks: "Code and data are available at: [https://github.com/wendyhuan/lifeexpectancy](https://github.com/wendyhuan/lifeexpectancy)."
date: 02 December 2024
date-format: long
abstract: "This paper analyze life expectancy and its determinants through 184 countries and make the prediction of people from different Income Group. Multiple Linear regression is used to deploying the life expectanc with gender, income and region. Predictions of life expectancy of people from different income group is made according to these essential predictors. The finding indicates that the life expectancy is tend to get higher as the economic class grows. And we predict that the average age of person from low, lower-middle, upper-middle and high are 62.35, 68.40, 73.38, 79.38. The result of this paper will help in suggesting a country which area should be given importance in order to efficiently improve the life expectancy of its population."
format: pdf
toc: true
pdf-engine: xelatex
geometry: landscape
toc-depth: 3
keep-tex: true
output:
  pdf_document:
    extra_dependencies: ["longtable"]
number-sections: true
bibliography: references.bib
---

@@

# Plot the percentage distribution
ggplot(life_expectancy_bins, aes(x = AgeRange, y = Percentage, fill = AgeRange)) +
  geom_bar(stat = "identity", color = "white", alpha = 0.8) +
  geom_text(aes(label = paste0(round(Percentage, 2), "%")), vjust = -0.5, size = 2.5) +
  labs(
    title = "Distribution of Life Expectancy by Age Range",
    x = "Life Expectancy (5-Year Ranges)",
    y = "Percentage"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

@@

# Create a model summary with median absolute deviation (mad) for statistics
modelsummary(
  list("Gaussian(Normal)" = model_developed),
  statistic = "mad",  # Median absolute deviation
  fmt = 2,     # Format results to 2 decimal places
  output = "kableExtra"
) %>%
  kable_styling(
    font_size = 7.5,
                full_width = FALSE,  # Keeps the table from spanning the full width
                position = "center", # Centers the table
                bootstrap_options = c("striped", "condensed"))

@@

The income group column and the region column was not given in the raw dataset. We created a new variable called "Income_Group" by mapping over country name to four income group ("lower_income", "lower_middle", "upper_middle", "high_income") according to the index given by World Bank Income Group. This is a key variable in predicting life expectancy since it is believed that life expectancy is correlated to the income situation. We as well made a mapping to the countries according to its continent and saved as "Region". This is the variable indicates the geographic context to the analysis. Countries are mapped into six continents ("Africa", "Asia", "Europe", "North America", "South America", "Oceania") one by one.

@@

```{r}
#| label: tbl-cleaned_average
#| tbl-cap: "Cleaned data of average life expectancy of 6 years by different countries."
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Create a nicely formatted kable table
bothg_average_life_expectancy %>%
  kable(
    col.names = c("Country", "Income Group", "Region", "Average Life Expectancy"),
    align = c("l", "l", "l", "l", "r"),
    digits = 2,
    output = "kableExtra"
) %>%
  kable_styling(font_size = 5.5,
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "condensed")
  ) 
```

@@

# Format the table using kable
results_df %>%
  head(100) %>%
  kable(col.names = c("Country", "Actual Life Expectancy", "Predicted Life Expectancy", "RMSE (Test)", "RMSE (Train)", "Indication"),
  align = c("l", "r", "r", "r", "r", "l"),
  digits = 2,
  output = "kableExtra"
) |>
  kable_styling(font_size = 8.5,
                full_width = FALSE,  # Keeps the table from spanning the full width
    bootstrap_options = c("striped", "condensed")) %>%
  column_spec(1, width = "0.35in") %>%
  column_spec(3:6, width = "1in")
